---
title: "palmer-penguins_markdown"
author: "E2"
date: "2026-01-03"
output: html_document
---

## Setting up the environment 
Notes: Loading tidyverse and Palmer penguins packages 
```{r loading packages}
library(tidyverse)
library(palmerpenguins)
```
  
## Explore the data set
Notes: Use tidyverse to perform exploratory data analysis  
  
### View dataset  
```{r view the entire tibble}
View(penguins)
```
  
### View the first 6 rows
```{r view first few rows}
head(penguins)
```
  
### Get column names
```{r get attributes }
colnames((penguins))
```
  
### Get statistical summary
```{r get statistical summary}
summary(penguins)
```
  
### View  the tibble's structure
```{r view tibble structure}
glimpse(penguins)
```

  
### Sort data by bill length in **Ascending** Order 
```{r sort by bill length ascending }
penguins %>% arrange(bill_length_mm)
```
  
### Sort data by bill length in **Descending** Order 
```{r sort by bill length descending}
penguins %>% arrange(-bill_length_mm)
``` 
  
### Find the average bill length in mm for penguins in each island
```{r calculate the average bill length by island}
penguins %>% group_by(island) %>%  drop_na() %>% 
  summarise(avg_bill_len_mm = mean(bill_length_mm)) 
```
  
### Find the average bill length in mm for each penguins species
```{r calculate the average bill length by species}
penguins %>% group_by(species) %>%  drop_na() %>% 
  summarise(avg_bill_len_mm = mean(bill_length_mm)) 
```
  
### Find the maximum bill length in mm for penguins in each island
```{r calculate the maximum bill length by island}
penguins %>% group_by(island) %>%  drop_na() %>% 
  summarise(max_bill_len_mm = max(bill_length_mm))
```
  
### Find the maximum body mass in kilograms for penguins in each island
```{r calculate the maximum body masss by island}
penguins %>% group_by(island) %>%  drop_na() %>% 
  summarise(max_body_mass_kg= max(body_mass_g) / 1000)
```
  
### Find the maximum body mass in kilograms for each penguin species
```{r calculate the maximum body masss by species}
penguins %>% group_by(species) %>%  drop_na() %>% 
  summarise(max_body_mass_kg= max(round(body_mass_g/1000,2)))
```  
  
### Create custom table of key statstical metrics
Notes: Find the maximum, average, standard deviation and minimum bill length in mm for penguins in each island and species
```{r create a custom summary}
penguins %>% group_by(island, species) %>%  drop_na() %>%
  summarise(min_bill_len_mm = min(bill_length_mm),
            avg_bill_len_mm = round(mean(bill_length_mm),2),
            max_bill_len_mm = max(bill_length_mm),
            std_bill_len_mm = sd(bill_length_mm))
```
  
## Filter the Palmer Penguins' data to discover details   
Notes: Create subsets of data to find information about specific demographics   
  
### Filter data for Adelie penguins
Note: This species has an equal number of males and females, they are the most popular and can be found on all three islands. 
```{r Get a statistical summary of the Adelie species}
adelie_penguins <- penguins %>%  
  filter(species == "Adelie") %>%  drop_na()
summary(adelie_penguins)
```
  
### Filter data for Gentoo penguins
Note: This species has more males than females and they can only be found on Biscoe island
```{r Get a statistical summary of the Gentoo species}
gentoo_penguins <- penguins %>%  
  filter(species == "Gentoo") %>%  drop_na()
summary(gentoo_penguins)
```
### Filter data for Chinstrap penguins
Note: This species has an equal number of males and females, they have the lowest population and all live on Dream island 
```{r Get a statistical summary of the Chinstrap species}
chinstrap_penguins <- penguins %>%  
  filter(species == "Chinstrap") %>%  drop_na()
summary(chinstrap_penguins)
```
### Filter by Dream island
Note: This island has two penguin species, males out number females by one.
```{r Get a statistical summary of the Dream island}
dream_island <- penguins %>% 
  filter(island == "Dream") %>% drop_na()
summary(dream_island)
```
  
### Filter data by Biscoe island
Note: This island has the most penguins, contains two penguin species and there are three more males than females 
```{r Get a statistical summary of the Biscoe island}
biscoe_island <- penguins %>% 
  filter(island == "Biscoe") %>%  drop_na()
summary(biscoe_island)
```

  
### Filter data by Torgersen island
Note: This island only contains one species of penguins and females out number males by one  
```{r Get a statistical summary of the Torgersen island}
torgersen_island <- penguins %>% 
  filter(island == "Torgersen") %>%  drop_na()
summary(torgersen_island)
```
  
## Visualise the Palmer Penguins' data 
Notes: Create different visuals to share data insights clearer  
  
### **Scatter plot**: flipper length vs  body mass 
Notes: Create a scatter plot to view flipper length and body mass relationship
```{r flipperlength vs Body mass }
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g)) + 
   geom_point()
```
  
### **Scatter plot(trend line)**: flipper length vs  body mass 
Notes: Create a scatter plot with a trend line to view flipper length and body mass relationship
```{r flipperlength vs Body mass ft Trendline }
ggplot(data = penguins, mapping = aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_smooth() + geom_point()
```

### **Scatter plot**: flipper length vs  body mass across species
Notes: Create a scatter plot to view flipper length and body mass relationship for each of the Penguin 🐧  species
```{r flipperlength vs Body mass color coded}
ggplot(data=penguins) +
  geom_point(mapping = aes(x=flipper_length_mm ,y=body_mass_g, colour = species))
```
### **Separate Scatter plots**: flipper length vs  body mass by species
Notes: Create scatter plots  to view flipper length and body mass relationship for each of the Penguin 🐧  species
```{r flipperlength vs Body mass separate }
ggplot(data=penguins) +
  geom_point(mapping = aes(x=flipper_length_mm ,y=body_mass_g, colour = species, shape = species)) +  facet_wrap(~species)
```

### **Separate Scatter plots(with trendlines)**: flipper length vs  body mass by species Trendlines
Notes: Create scatter plots  to view flipper length and body mass relationship for each of the Penguin 🐧  species and trend lines for each
```{r flipperlength vs Body mass separate with Trendlines }
ggplot(data=penguins, mapping = aes(x=flipper_length_mm ,y=body_mass_g, colour = species, shape = species))  +
  geom_smooth() + geom_point() + facet_wrap(~species)
```

### **Final Scatter plot**: flipper length vs  body with text labels
Notes: Create a colorful scatter plot to view flipper length and body mass relationship for each of the Penguin 🐧  species title, subtitle,caption, and annotation and trend line 
```{r flipperlength vs Body mass with labels }
ggplot(data=penguins) +
  geom_smooth(mapping = aes(x=flipper_length_mm,y=body_mass_g,colour= "violet")) +
  geom_point(mapping = aes(x=flipper_length_mm ,y=body_mass_g, colour = species)) +
  labs(title = "Palmer Penguins: Body mass vs flipper length",
       subtitle = "Sample of three species", 
       caption = "Data collected by Dr. Kristen Gorman") + 
  annotate("text",x= 220, y= 4110,label="Gentoos are the largest",
           colour= "navyblue", font_face="bold",size=4.3, angle= 30)
```
  
### **Adelie Penguins Bar plot**: sex on  each island  
```{r Adelie demographics}
ggplot(data = adelie_penguins) +
  geom_bar(mapping = aes(x = island,fill = sex)) +
  labs(title = "Demographics of Adeline Penguins ")
```
  
### **Gentoo Penguins Bar plot**: sex on  each island  
```{r Gentoo demographics}
ggplot(data = gentoo_penguins) +
  geom_bar(mapping = aes(x = island,fill = sex)) +
  labs(title = "Demographics of Gentoo Penguins ")
```
  
### **Chinstrap Penguins Bar plot**: sex on  each island  
```{r Chinstrap demographics}
ggplot(data = chinstrap_penguins) +
  geom_bar(mapping = aes(x = island,fill = sex)) +
  labs(title = "Demographics of Chinstrap Penguins ")
```
  
### **Biscoe island Bar plot**: species and sex  
```{r Biscoe island Demographics }
ggplot(data = biscoe_island) + 
  geom_bar(mapping = aes(x = species, fill = sex))
```
  
### **Dream island Bar plot**: species and sex  
```{r Dream island Demographics}
ggplot(data = dream_island) + 
  geom_bar(mapping = aes(x = species, fill = sex))
```
  
### **Torgersen island Bar plot**: species and sex  
```{r Torgersen island Demographics}
ggplot(data = torgersen_island) + 
  geom_bar(mapping = aes(x = species, fill = sex))
```
